---
title: "Assignment 6: Data Scraping"
author: "Ashlee Warren"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 loading tidyverse, rvest, and here
library(tidyverse)
library(rvest)
library(here)

#verifying my working directory
here()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
Durham_LWSP_Website <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')

Durham_LWSP_Website
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 creating the water system name variable
water_system_name <- Durham_LWSP_Website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()

#creating the PWSID variable
PWSID <- Durham_LWSP_Website %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()

#creating the ownership variable
ownership <- Durham_LWSP_Website %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()

#creating the maximum daily use for each month of 2024 variable
maximum_daily_use_monthly <- Durham_LWSP_Website %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 creating a dataframe with the columns of the four variables I just scraped and a row for the months
Durham_2024_LWSP <- data.frame("water_system_name" = as.character(water_system_name),
                               "PWSID" = as.character(PWSID), 
                               "ownership" = as.character(ownership),
                               "maximum_daily_use_monthly" = as.numeric(maximum_daily_use_monthly))

#adding a date column to the dataframe
Durham_2024_LWSP <- Durham_2024_LWSP %>% 
  mutate(Date = my(paste(c("Jan", "May", "Sept", "Feb", "June", "Oct", "Mar", "Jul",
                           "Nov", "Apr", "Aug", "Dec"), 2024))) %>% 
  mutate(Year = year(Date))

#viewing the dataframe and commenting it out after
#view(Durham_2024_LWSP)

#5 creating a line plot for the maxium daily use values for each month in 2024
ggplot(Durham_2024_LWSP, aes(x=Date, y=maximum_daily_use_monthly)) +
  geom_line() +
  labs(title = "Maximum Day Use of Water in Durham, NC in 2024",
       x= "Date",
       y="Maximum Day Use (MGD)")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6 creating a scraping function with PWSID and year
scrape.it.LWSP <- function(PWSID, Year){
  #retrieving the LWSP website contents
  NC_LWSP_Website <- read_html(paste0(
'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',PWSID,'&year=',Year))
  #setting the element variables
  PWSID_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  Year_tag <- '.fancy-table:nth-child(64) tr:nth-child(1) th:nth-child(2)'
  #scraping the data items
  PWSID_website <- NC_LWSP_Website %>% 
    html_nodes(PWSID_tag) %>% 
    html_text()
  
  Year_website <- NC_LWSP_Website %>% 
    html_nodes(Year_tag) %>% 
    html_text()
  
  #creating the dataframe
  NC_LWSP <- data.frame("water_system_name" = as.character(water_system_name),
                               "ownership" = as.character(ownership),
                               "maximum_daily_use_monthly" = 
                                as.numeric(maximum_daily_use_monthly)) %>% 
                mutate(PWSID = !!PWSID,
                       Year = !!Year,
                       Month = m(c("Jan", "May", "Sept", "Feb", "June", "Oct", "Mar", 
                                   "Jul", "Nov", "Apr", "Aug", "Dec")))
  
  #returns the dataframe as the function's output
  return(NC_LWSP)
  }

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 using my scrape.it function to extract the max daily use for Durham for each 
#month in 2020 and return it in a dataframe
Durham_2020_LWSP <- scrape.it.LWSP('03-32-010', 2020)

#viewing the structure of my dataframe
glimpse(Durham_2020_LWSP)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 


```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 


```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
>
