---
title: "Assignment 4: Data Wrangling"
author: "Ashlee Warren"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  markdown: 
    wrap: 72
---

## OVERVIEW

In this assignment, you will build a reproducible workflow that cleans
four EPA air-quality files, combines them, reshapes the result, and
produces grouped summaries. Date conversion is a brief review from
Module 3; the emphasis is on data wrangling.

## Directions

1.  Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.
6.  Ensure that code in code chunks does not extend off the page in the
    PDF.

## Set up your session

### Questions 1–2

1a. Load the `tidyverse` and `here` packages into your session.

1b. Check your working directory.

1c. Read all four raw files associated with the EPA Air dataset. See the
README file for information about the variables and measurements.

2.  Add the appropriate code to reveal the dimensions of the four
    datasets.

```{r setup questions 1-2, message = FALSE}
#1a Loading the tidyverse and here packages
library(tidyverse)
library(here)

#1b Checking the location of my working directory
here()

#1c Adding the four raw data files for the EPA Air dataset
EPAair.O3.NC2024 <- read.csv(
  here("Data/Raw/EPAair_O3_NC2024_raw.csv")
)

EPAair.O3.NC2025 <- read.csv(
  here("Data/Raw/EPAair_O3_NC2025_raw.csv")
)

EPAair.PM25.NC2024 <- read.csv(
  here("Data/Raw/EPAair_PM25_NC2024_raw.csv")
)

EPAair.PM25.NC2025 <- read.csv(
  here("Data/Raw/EPAair_PM25_NC2025_raw.csv")
)

#I am also reviewing the metadata for each to understand the datasets and variables.

#2 #Adding code to find the dimensions of each dataset
dim(EPAair.O3.NC2024)
#EPAair.O3.NC2024 has 9,519 observations of 21 variables

dim(EPAair.O3.NC2025)
#EPAair.O3.NC2025 has 9,466 observations of 21 variables

dim(EPAair.PM25.NC2024)
#EPAair.PM25.NC2024 has 9,752 observations of 22 variables

dim(EPAair.PM25.NC2025)
#EPAair.PM25.NC2025 has 10,184 observations of 22 variables
```

> All four datasets should have the same number of columns (PM2.5 has
> one extra column), but unique record counts (rows). Do your datasets
> follow this pattern? Yes, my O3 datasets both have 21 columns, and my
> PM25 datasets have 22 columns. Each of the four datasets have unique
> number of rows.

\pagebreak
## Wrangle individual datasets to create processed files.

### Questions 3–5

3.  Change the Date columns to be date objects.

4.  Select the following columns: DDate, Daily.AQI.Value,
    Local.Site.Name, AQS.Parameter.Description, County, Site.Latitude,
    Site.Longitude

5.  For the PM2.5 datasets, fill all cells in AQS.Parameter.Description
    with "PM2.5" (all cells in this column should be identical).

```{r questions 3 thru 5}
#3 Changing the date columns in all four datasets from character objects to date objects

EPAair.O3.NC2024$Date <- mdy(EPAair.O3.NC2024$Date)

EPAair.O3.NC2025$Date <- mdy(EPAair.O3.NC2025$Date)

EPAair.PM25.NC2024$Date <- mdy(EPAair.PM25.NC2024$Date)

EPAair.PM25.NC2025$Date <- mdy(EPAair.PM25.NC2025$Date)

#4 Selecting these columns from all four datasets: Date, Daily.AQI.Value, 
#Local.Site.Name, AQS.Parameter.Description, County, Site.Latitude, 
#Site.Longitude

EPAair.O3.NC2024.processed <- select(EPAair.O3.NC2024, Date, Daily.AQI.Value,
                                     Local.Site.Name, AQS.Parameter.Description, 
                                     County:Site.Longitude)

EPAair.O3.NC2025.processed <- select(EPAair.O3.NC2025, Date, Daily.AQI.Value, 
                                     Local.Site.Name, AQS.Parameter.Description, 
                                     County:Site.Longitude)

EPAair.PM25.NC2024.columnsubset <- select(EPAair.PM25.NC2024, Date, Daily.AQI.Value, 
                                          Local.Site.Name, AQS.Parameter.Description, 
                                          County:Site.Longitude)

EPAair.PM25.NC2025.columnsubset <- select(EPAair.PM25.NC2025, Date, Daily.AQI.Value, 
                                          Local.Site.Name, AQS.Parameter.Description, 
                                          County:Site.Longitude)

#5 First, I want to understand what is currently in AQS.Paramenter.Description column 
#of the PM25 datasets, so I am viewing the 2024 dataset and then commenting it out. 
#Next, I'm changing all cells in the AQS.Parameter.Description column to "PM2.5". I 
#am then viewing both datasets to ensure that it worked and commenting out view.

#view(EPAair.PM25.NC2024.columnsubset)

EPAair.PM25.NC2024.processed <- mutate(EPAair.PM25.NC2024.columnsubset, 
                                       AQS.Parameter.Description = "PM2.5")

#view(EPAair.PM25.NC2024.processed)

EPAair.PM25.NC2025.processed <- mutate(EPAair.PM25.NC2025.columnsubset, 
                                       AQS.Parameter.Description = "PM2.5")

#view(EPAair.PM25.NC2025.processed)

```

### Question 6

6.  Save all four processed datasets in the Processed folder. Use the
    same file names as the raw files but replace "raw" with "processed".

```{r question 6}
#6 Saving all four processed datasets in the Processed folder and replacing "raw" with 
#"processed" in the file name.

write.csv(
  EPAair.O3.NC2024.processed,
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_NC2024_processed.csv")
)

write.csv(
  EPAair.O3.NC2025.processed,
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_NC2025_processed.csv")
)

write.csv(
  EPAair.PM25.NC2024.processed,
  row.names = FALSE,
  file = here("Data/Processed/EPAair_PM25_NC2024_processed.csv")
)

write.csv(
  EPAair.PM25.NC2025.processed,
  row.names = FALSE,
  file = here("Data/Processed/EPAair_PM25_NC2025_processed.csv")
)
```

## Combine datasets

### Question 7

7.  Stack the four datasets with `bind_rows()`. Check that their column
    names are identical first.

```{r question 7}
#7 Checking the column names for each dataset

colnames(EPAair.O3.NC2024.processed)
colnames(EPAair.O3.NC2025.processed)
colnames(EPAair.PM25.NC2024.processed)
colnames(EPAair.PM25.NC2025.processed)

#Confirmed that the column names are the same and am binding the rows together into on dataframe
EPAair.O3.PM25.2024.2025 <- bind_rows(EPAair.O3.NC2024.processed, 
                                      EPAair.O3.NC2025.processed, 
                                      EPAair.PM25.NC2024.processed, 
                                      EPAair.PM25.NC2025.processed)
```

### Question 8

8.  Use one pipeline (`%>%`) to create a daily site-level dataset that
    meets the following conditions:

-   Include only sites that the four data frames have in common:

"Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",\
"Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West
Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri.
Center", "Bryson City", "Millbrook School"

(the function `intersect` can figure out common factor levels - but it
will include sites with missing site information, which you don't
want...)

-   Some sites have multiple measurements per day. Group by date, site
    name, AQS parameter, and county, then calculate the mean AQI value,
    latitude, and longitude. Handle missing values explicitly in each
    mean.

-   Add numeric `Month` and `Year` columns from `Date`. Check if results
    are integers.

-   Hint: the dimensions of this dataset should be 15,012 × 9.

```{r question 8}
#8 I am using a pipe to create a daily site-level dataset from my combined dataframe. 
#First, I am using the filter argument to filter the Local Site Name column to only 
#include the sites that all four datasets have in common. Second, I am using the 
#group_by function to group by date, site name, AQS parameter, and county. Next, I 
#am using the summarize function to create new columns that have the mean for the 
#Daily AQI value, the Site Latitude, and the Site Longitude and preserving the NAs 
#while asking the mean to ignore them. I'm ending that function with .groups = "drop" 
#to remove the grouping after the summary is complete. Next, I am using the mutate 
#argument to add columns for the month and year and using as.integer to ensure that 
#they are stored as integers rather than numerics.

EPAair.O3.PM25.2024.2025.DailySiteLevel <-
  EPAair.O3.PM25.2024.2025 %>% 
  filter(Local.Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", 
           "Hattie Avenue", "Clemmons Middle", "Mendenhall School", 
           "Frying Pan Mountain", "West Johnston Co.", 
           "Garinger High School", "Castle Hayne", "Pitt Agri. Center", 
           "Bryson City", "Millbrook School")) %>% 
  group_by(Date, Local.Site.Name, AQS.Parameter.Description, County) %>% 
  summarize(
    Mean.Daily.AQI = mean(Daily.AQI.Value, na.rm = TRUE),
    Mean.Site.Latitude = mean(Site.Latitude, na.rm = TRUE),
    Mean.Site.Longitude = mean(Site.Longitude, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(Month = as.integer(month(Date)),
         Year = as.integer(year(Date)))
  
```

### Questions 9–10

9.  Use `pivot_wider()` so that ozone and PM2.5 AQI values occupy
    separate columns. Each location-date combination should occupy one
    row.

10. Call up the dimensions of your new tidy dataset.

```{r questions 9-10}
#9 I'm using the pivot_wider so that ozone and PM2.5 values have separate columns.

EPAair.O3.PM25.2024.2025.DailySiteLevel.wide <-
  EPAair.O3.PM25.2024.2025.DailySiteLevel %>% 
  pivot_wider(
    names_from = AQS.Parameter.Description,
    values_from = Mean.Daily.AQI
  )

#view(EPAair.O3.PM25.2024.2025.DailySiteLevel.wide) checked to make sure the new dataframe
#looked as intended and am not commenting it out.

#10 Using dim to get the dimensions of my wide dataframe.
dim(EPAair.O3.PM25.2024.2025.DailySiteLevel.wide)

#The wide dataframe has 9,037 observations of 9 variables.

```

### Question 11

11. Save your processed dataset with the following file name:
    "EPAair_O3_PM25_NC2425_Processed.csv"

```{r question 11}
#11 Saving the wide dataframe as a new processed csv file

write.csv(
  EPAair.O3.PM25.2024.2025.DailySiteLevel.wide,
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_PM25_NC2425_Processed")
)


```

## Generate summary tables

### Questions 12–13

12. Create a monthly summary data frame grouped by site, month, and
    year. Calculate mean ozone and PM2.5 AQI for each group, handling
    missing values explicitly. Then use `drop_na()` to remove rows only
    when mean **ozone** is missing. Mean PM2.5 may remain missing.

13. Call up the dimensions of the summary dataset.
\pagebreak
```{r questions 12 and 13}
#12 Creating a monthly summary dataframe. First, I'm using group_by to group by the site, 
#month, and year. Next, I'm calculating the mean O3 and PM2.5, ignoring the NAs, and 
#removing the groupings after the summary is complete. Finally, I'm removing the rows 
#with mean O3 is missing.

EPAair.O3.PM25.NC2425.MonthlySummary <-
  EPAair.O3.PM25.2024.2025.DailySiteLevel.wide %>% 
  group_by(Local.Site.Name, Month, Year) %>% 
  summarize(mean.PM2.5 = mean(PM2.5, na.rm = TRUE),
            mean.O3 = mean(Ozone, na.rm = TRUE),
            .groups = "drop") %>% 
  drop_na(mean.O3)


#13 Finding the dimensions of my monthly summary dataframe
dim(EPAair.O3.PM25.NC2425.MonthlySummary)

#New dimensions include 238 observations of 5 variables. I'm comparing this finding using 
#drop_na with na.omit by creating a second monthly summary dataframe using that argument.

EPAair.O3.PM25.NC2425.MonthlySummary.Omit <-
  EPAair.O3.PM25.2024.2025.DailySiteLevel.wide %>% 
  group_by(Local.Site.Name, Month, Year) %>% 
  summarize(mean.PM2.5 = mean(PM2.5, na.rm = TRUE),
            mean.O3 = mean(Ozone, na.rm = TRUE),
            .groups = "drop") %>% 
  na.omit(mean.O3)

#alternative dataframe is created, and I'm checking the dimensions using dim.

dim(EPAair.O3.PM25.NC2425.MonthlySummary.Omit)

#The dimensions for this one are 234 observations of 5 variables.
```

### Question 14

14. Why is `drop_na(mean_ozone)` more appropriate here than `na.omit()`?
    Test both approaches and compare the dimensions of the resulting
    summary data frames.

> Answer: drop_na resulted in 238 rows while na.omit only resulted in
> 232 rows. This indicates that na.omit was less targeted and deleted
> six rows that I were needed.
