---
title: "Assignment 3: Data Exploration"
author: "Ashlee Warren"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Exploration M3: Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2016-26_raw.csv). Name these datasets "Neonics" and "Litter", respectively.

```{r setup,  message = FALSE}
#load packages
library(tidyverse)
library(lubridate)
library(here)

#Checking my current working directory
here()

#Importing my datasets
Neonics <- read.csv(
  file = here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.txt"),
  sep = "|"
  )

Litter <- read.csv(
file = here("Data/Raw/NEON_NIWO_Litter_massdata_2016-26_raw.csv")
)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: I think we'd be interested in the ecotoxicology of neonicotinoids for two reasons. First, I think we would want to understand their effectiveness in controling the agriculture pests they are designed to address. Second, I understand that neonicotinoids can harm honeybees and other pollinators, and we would want to better understand that impact on those insects.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: I think this is of interest as wildfires are becoming more intense. Litter and woody debris are one contributing factor, and it would be helful to study that connection and monitor the amount of debris. Litter and woody debris are also important in an ecosystem's nutrient cycling and provide important wildlife habitat. Monitoring it coud help us understand the health of that ecosystem.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Sampling occurs only in Tower plots, and their locations were randomly selected "within the 90% flux footprint of the primary and secondary airsheds."
 2. Sampling at forested tower airshed sites occurs "in up to two randomly selected subplots in each of 20 40m x 40m plots" while low-saturated vegegation sites have "up to 30 20m x
20m plots." The number of plots can sometimes be lower if space doesn't allow.
 3. For every 400 square meters plot area, there is one elevated trap and/or one ground trap. Ground traps are sampled once per year. Elevated trap sampling is dependent upon vegetation with sampling occuring once every 2 weeks in deciduious forests and once every 1 to 2 months in coniferous forests. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
#Finding the dimensions of the Neonics dataset
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}
#Using the summary function on the Effect column to determine the most common effects
Summary(Neonics$Effect)

#This only told me the length, class, and mode information rather than the actual effects. I'm converting the Effect column data to a factor and then using the summary function.
Neonics$Effect <- factor(Neonics$Effect)
summary(Neonics$Effect)

```

> Answer: The four most common effects are Mortality, Population, Genetics, and Behavior. These are particularly interesting because they are high impact effects. The insect dying and changes in population, genetics, and behavior all have significant meaning for that species and potentially, the ecosystem.

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}
#Using the factor function first to convert the text into data that is sortable and then using the summary function to determine the six mostly commonly studied species
Neonics$X.Species.Common.Name <-factor(Neonics$X.Species.Common.Name)
summary(Neonics$X.Species.Common.Name)
summary(Neonics$X.Species.Common.Name, maxsum = 6)
summary(Neonics$X.Species.Common.Name, maxsum = 7)

#After using the summary feature initially, I tried out the maxsum addition. When I selected 6, it grouped all of the other species as the sixth item on the list. I then ran it for seven, and it gave me the six most studied species with the seventh listing as all others.
```

> Answer: The six most commonly studied species are honey bee, Carniolan honey bee, parasitic wasp, bumble bee, Italian honeybee, and Asiatic honey bee. All of these species are pollinators, which would make them of great interest to study.

8.  Concentrations are always a numeric value. What is the class of `Conc.1.Mean..Author..` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
#Viewing the dataframe to better understand the data in Conc.1.Mean..Author.. and then checking the class
view(Neonics)
class(Neonics$X.Conc.1.Mean..Author..)
```

> Answer: The class of Conc.1.Mean..Author.. is character. I believe this may be the case because some of the numbers have a / following the number.

## Explore your data graphically (Neonics)

9. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}


```

> Answer:

## Explore your data (Litter)

10. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}


```

11. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}


```

> Answer:

12. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}


```


What type(s) of litter tend to have the highest biomass at these sites?

> Answer:

13. Write a short data-exploration summary for Neonics.

> - **One key characteristic:** 
> - **One data-quality issue:** 
> - **One interesting pattern:** 
> - **One question for further analysis:** 

14. Write a short data-exploration summary for Litter.

> - **One key characteristic:** 
> - **One data-quality issue:** 
> - **One interesting pattern:** 
> - **One question for further analysis:** 
